Human Factors Influencing Visual Statistical Inference

### 2013-06-15 sociology_paper.tex
reviewd upto location effect

\subsection{Choice of Visual Test Statistic}  The visual test statistic should be highly associated with the hypothesis under investigation. To achieve this purposes it is very important to decide which plot type and plot features should be adopted.  In a linear regression setting, the visual test statistics are presented in \citep{majumder:2013} for some common hypothesis testing.  It is also observed that a scatterplot may do a better job than a box plot when using as a visual test statistic for regression parameters.  Some of the effective features of visual test statistics are discussed in \citep{heike:2012} including plot type, color and shape of the plots.  \cite{niladri:2012} presents some \textit{distance measure} to determine how a plot may be different from each other.

Table \ref{tbl:visual_stat} shows the visual test statistics used in this study. For both experiment 2 and 3 a scatterplot is used but a regression line fitted through the points is overlaid for experiment 2. It is easier to spot the slope of the line compared to just the scatterplot itself. But it appears that people are better in noticing the unusual pattern in the data which is some contamination purposefully added for experiment 3. It is observed that for experiment 3 the performance of the observers are much better compared to experiment 2 \citep{majumder:2013}.

@@

\subsection{Question that Human Observer Answers} The researcher knows about the underlying hypothesis but the observer does not necessarily know the underlying details of the lineup. So, the researcher needs to ask a question observer is to answer while evaluating the lineup. This question should provide the observer a little clue so that the answer reflects the hypothesized patterns in the actual plot. For example Table \ref{tbl:visual_stat} shows the questions asked for the simulation experiments done by  \citet{majumder:2013}. Notice that for case 1 if the observer can identify the actual plot in the lineup that should indicate that the plot chosen has the most vertical difference between groups A and B which is exactly what the researchers intend to examine. Similarly for case 2, a correct evaluation would indicate that the slope is different than the slope that may show up just from randomness.

These questions are very crucial for the power of visual test. They help observer think in a meaningful direction. Notice that there may be unnecessary patterns in the actual data plot which may not necessarily indicate the existence of the significant signal in the plot. These question help observer not to be misguided by those patterns. To review further on this \citet{majumder:2013} have also asked why the observers choose a specific plot.

@@

\subsection{Demographics of Observer}  Some of the demographics of the observer such as age, gender, education level and geographical location may have effect on how an observer would examine the lineup. This may produce some variability in the performance of the observer. For example, a high school student may not respond same as a well educated person while evaluating a lineup. There may be variation in different age groups as well. Recruiting subjects with a well variety of age groups is be necessary to study this variability.

@@

\subsection{Learning Trend of the Observer} When an observer evaluates a lineup he or she may learn something from the experiences of the earlier observations. If the same observer is shown another lineup the learning from previous evaluation may help. The more evaluations an observer makes the more skillful the observer may become. This learning trend may or may not be significant overall but it could influence the performance of the observer in some way.

Learning may occur in two different ways. An observer can become more trained on the lineup structure and the pattern in the plots shown. This could be learning in proportion of correct evaluation. Or, he or she may become efficient in responding faster in the later evaluations. This could be observed as the time taken for evaluating a lineup. This paper investigates both of these learning trends with multiple simulation experiments.

When an observer evaluate a lineup for the first time, he or she has to read through many instructions and become familiar with the environment of the experiment. We expect that this would require much more time compared to the rest of the evaluations. Time taken for the second or successive attempt would be much lower than the time taken for the first lineup. We attribute this to the adaptation to the experiment environment, not a learning of evaluating a lineup.

\subsection{Location of Actual Plot in the Lineup} The actual data plot is placed in a random spot in a lineup. While evaluating the lineup some people may start looking from some specific part of the lineup. With the help of eye-tracking equipment \citet{zhao:2012} track the observers' eyes to see how they go through the plots in a lineup to come to their answers. The results suggest that people have particular methods of reading lineups. Some people read lineups from left to write direction while some read from upward to downward. Some people start looking from the center of the lineup while others start from the top left corner. In the earlier phase of the exercise, the observer tend to scan the plots and start comparing plots to make a final decision. Beside right-left or up-down directions observer show some diagonal movement too.

@@

It may be possible that some part of the lineup is little explored or not scanned by the observer if he or she feels that the actual plot is found before even seriously scanning the whole lineup. In those scenarios the location where the observer first start scanning is important. Placing the actual data plot in that location may yield different results.

>>>>

\subsection{Choice of Visual Test Statistic}  The visual test statistic should be highly associated with the hypothesis under investigation. To achieve this purpose it is very important to decide which plot type and plot features should be adopted.  In a linear regression setting, the visual test statistics are presented in \citep{majumder:2013} for some common hypothesis testing.  It is also observed that a scatterplot may do a better job than a box plot when using as a visual test statistic for regression parameters.  Some of the effective features of visual test statistics are discussed in \citep{heike:2012} including plot type, color and shape of the plots.  \cite{niladri:2012} presents some \textit{distance measure} to determine how a plot may be different from each other.

Table \ref{tbl:visual_stat} shows the visual test statistics used in this study. For both experiments 2 and 3 a scatterplot is used but a regression line fitted through the points is overlaid for experiment 2. It is easier to spot the slope of the line compared to just the scatterplot itself. But it appears that people are better in noticing the unusual pattern in the data which is some contamination purposefully added for experiment 3. For experiment 3 the performance of the observers was much better compared to experiment 2 \citep{majumder:2013}.

@@

\subsection{Question that Human Observer Answers} The researcher knows about the underlying hypothesis but the observer does not necessarily know the underlying details of the lineup. So, the researcher needs to ask a question to the observer to answer while evaluating the lineup. This question should provide the observer a little clue so that the answer reflects the hypothesized patterns in the actual plot. For example Table \ref{tbl:visual_stat} shows the questions asked for the simulation experiments considered for this study. Notice that for case 1 if the observer can identify the actual plot in the lineup that should indicate that the plot chosen has the most vertical difference between groups A and B which is exactly what the researchers intend to examine. Similarly for case 2, a correct evaluation would indicate that the slope is different than the slope that may show up just from randomness.

These questions are very crucial for the power of visual test. They help observer think in a meaningful direction. Notice that there may be unnecessary patterns in the actual data plot which may not necessarily indicate the existence of the significant signal in the plot. These questions help observer not to be misguided by those patterns. To follow up further on this \citet{majumder:2013} also collected reasons of choice made by the observers and noticed that Type-III error may occur where the hypothesis may be correctly rejected but for a wrong reason.

@@

\subsection{Demographics of the Observer}  Some of the demographics of the observer such as age, gender, education level and geographical location may have effect on how an observer would examine the lineup. This may produce some variability in the performance of the observer. For example, a high school student may not respond same as a well educated person while evaluating a lineup. There may be variations in different age groups as well. Investigation on subjects with a well variety of age groups is necessary to study this variability.

@@

\subsection{Learning Trend of the Observer} When an observer evaluates a lineup he or she may learn something from the experiences of the earlier observations. If the same observer is shown another lineup the learning from previous evaluation may help. The more evaluations an observer makes the more skillful the observer may become. This learning trend may or may not be significant overall but it may influence the performance of the observer in some way.

Learning may occur in two different ways. An observer can become more trained on the lineup structure and the pattern in the plots shown. This can be learning in proportion of correct evaluation. Or, he or she may become efficient in responding faster in the later evaluations. This can be observed as the time taken for evaluating a lineup. This paper investigates both of these learning trends with multiple simulation experiments.

When an observer evaluate a lineup for the first time, he or she has to read through many instructions and become familiar with the environment of the experiment. We expect that this will require much more time compared to the rest of the evaluations. Time taken for the second or successive attempt should be much lower than the time taken for the first lineup. We attribute this to the adaptation to the experiment environment, not a learning of evaluating a lineup.

\subsection{Location of Actual Plot in the Lineup} The actual data plot is placed in a random spot in a lineup. While evaluating the lineup some people may start looking from some specific part of the lineup. With the help of eye-tracking equipments \citet{zhao:2012} tracked the observers' eyes to see how they went through the plots in a lineup to come to their answers. The results suggest that people have particular methods of reading lineups. Some people read lineups from left to right direction while some read from upward to downward. Some people start looking from the center of the lineup while others start from the top left corner. In the earlier phase of the exercise, the observer tend to scan the plots and start comparing plots to make a final decision. Beside right-left or up-down directions observers show some diagonal movement too.

@@

It may be possible that some part of the lineup is less explored or even never scanned by the observer if he or she feels that the actual plot is found before even seriously scanning the whole lineup. In those scenarios the location where the observer first start scanning is important. Placing the actual data plot in that location may yield different results.
